---
title: "Visualization in R with ggplot2"
author: "John Little"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
## Code Repository

Download code for this workshop

https://github.com/libjohn/workshop_ggplot2


---
## Example 

```{r message=FALSE, warning=FALSE, include=FALSE}
library(babynames)
library(tidyverse)
```

```{r echo=FALSE, fig.height=6, fig.width=10}
starwars %>%
  drop_na(mass, height, birth_year, gender) %>%
  filter(mass < 400,
         species %in% c("Human", "Droid", "Gungan",
                        "Mirialan", "Wookiee")) %>%
  ggplot(aes(x = birth_year, y = height)) +
  geom_point(
    aes(color = gender, size = mass, shape = species),
        alpha = 0.7) +
  labs(title = "Star Wars species height over year",
       caption = "source:  dplyr::starwars")
```

.footnote[.tiny[Demonstrates some affordances of the ggplot2 package. Not intended as an example of visualization best practices.]]

---
class: middle
## Attribution

Adapted in whole or in part based on the [Visualize Data with ggplot2](https://github.com/rstudio/master-the-tidyverse/blob/master/pdfs/01-Visualize-Data.pdf) slides by Garrett Grolemund at RStudio.  CC BY Garrett Grolemund, RStudio

[CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/)  John Little

---
class: middle
## ggplot2 template


```r
ggplot(data = ---, mapping = aes(x = ---, y = ---)) +
  geom_----()

```

This assumes you loaded the `ggplot2` library package
```{r message=FALSE, warning=FALSE, INCLUDE=TRUE}
library(tidyverse)
# library(ggplot2)  # ggplot2 is included in tidyverse
```


---
class: middle, duke-orange
## mpg 

An on-board (`ggplot2`) highway mileage dataset similar to `mtcars`

```r
mpg
```

What relationship do you expect to see between engine
size (`displ`) and mileage (`hwy`)?

If time allows, familiarize yourself with the `mpg` dataset

```{r echo=FALSE}
countdown::countdown(1)
```


---
class: middle, duke-orange
## Your Turn 1

Open `01_exercises.Rmd` in RStudio

Inside the _Your Turn 1_ code chunk, in your `01.exercises.Rmd` notebook, make a graph by typing the following code.


```r
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

Pay **strict attention** to _spelling, capitalization_, and _parentheses_!


```{r echo=FALSE}
countdown::countdown(1)
```

---
class: middle
```{r echo=FALSE, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point()
```

---
class: middle
```{r echo=FALSE, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

---
class: middle

### Template
```r
*ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Alternative, tidy, concise
.content-box-grey[
```r
*mpg %>% 
*  ggplot() + 
   geom_point(aes(displ, hwy))  
```
]

1. "Initialize" a plot with ggplot()
2. Add layers with [geom_____ functions](https://ggplot2.tidyverse.org/reference/index.html#section-layer-geoms)


---
## Mappings

```r
mpg %>%
  ggplot() +
* geom_point(aes(displ, hwy))
```

```{r echo=FALSE, fig.height=5, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  stat_ellipse(data = . %>% filter(class == "2seater"), geom = "polygon", alpha = 0.3)
```
 
---
## Aesthetics

- Color
- Size
- Line type
- Opacity
- Shape

### Map data cell values to aesthetics

.center[`aes(color = class)`]

Visual Space | &nbsp; | Data Space
---|---|---
`color` | &nbsp; | `mpg$class`
.red[Red] | $\Leftrightarrow$ | 2 seater
.blue[Blue] | $\Leftrightarrow$ | pickup
... | $\Leftrightarrow$ | ...
.green[Green] | $\Leftrightarrow$ | midsize


---
```r
mpg %>% 
  ggplot(aes(displ, hwy)) +
* geom_point(aes(color = class))
```

```{r echo=FALSE, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  stat_ellipse(data = . %>% filter(class == "2seater"))
```


---
class: middle


```r
ggplot(mpg) + geom_point(aes(x = displ, y = hwy, color = class))
       ... ... ...                               size = class))
       ... ... ...                               shape = class))
       ... ... ...                               alpha = class))
```


---
class: duke-orange
## Your Turn 2

Re-focus in `01_exercises.Rmd`  

### Experiment

1. Add color, size, alpha, and shape arguments to your `aes()` function
1. Map different aesthetics to different variables
1. How do continuous and discrete variables affect the aesthetic mappings


```{r echo=FALSE}
countdown::countdown(5)
```

---
class: middle, duke-softblue
## Discrete v Continuous

Aesthetics arguments (`color`, `size`, `shape`, etc.) are affected by mode and type of data

Is your data 

- Discrete (Categorical: ordinal, nominal) (See:  [`library(forcats)`](https://forcats.tidyverse.org/))
- Continuous (Numeric)

NOTE:  `shape` argument will not work with continuous data

.footnote[See:  _R4DS_ [Important types of atomic vectors](https://r4ds.had.co.nz/vectors.html#important-types-of-atomic-vector) &  _R4DS_ [Factors](https://r4ds.had.co.nz/factors.html)]

---
## Mapping v Setting

```{r echo=TRUE, fig.height=6, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point(color = "blue") 
```


---
class: middle

.pull-left[
```{r echo=TRUE}
mpg %>%
  ggplot(aes(displ, hwy)) +
 geom_point(aes(color = class)) #<< 
```

]

.pull-right[

```{r echo=TRUE}
mpg %>% 
  ggplot(aes(displ, hwy)) +
{{ geom_point(color = "blue") }}
```

]

---

## Geoms

[Geometric Objects](https://ggplot2.tidyverse.org/reference/index.html#section-layer-geoms)

.pull-left[

```{r}
mpg %>% 
  ggplot(aes(displ, hwy)) +
 {{ geom_point()  }}
``` 

]

.pull-right[


```{r message=FALSE, warning=FALSE}
mpg %>% 
  ggplot(aes(displ, hwy)) +
{{ geom_smooth()  }}
```

]


---
## A few common geoms

Type | Geom
---|---
Bar graph | `geom_bar()  geom_col()`
Histogram | `geom_hist`
Scatter plot | `geom_point()`
Line graph | `geom_line()`
Box plot | `geom_boxplot()`
Density | `geom_density()  geom_violin()`
Heat map | `geom_heatmap()`
Mapping | `geom_sf()`
Regression line | `geom_smooth()`

.center[[More Geometric Objects](https://ggplot2.tidyverse.org/reference/index.html#section-layer-geoms)]

.footnote[See Also:  RStudio Menubar > Help > Cheatsheets > Visualization with ggplot2]

---
class: duke-orange
## Your Turn

Re-focus in `01_exercises.Rmd`  

See Your Turn 3 through 7

Answer can be found in `01_exercises_answers.Rmd`

```{r echo=FALSE}
countdown::countdown(10)
```
---
## Global v local

```{r echo=FALSE, fig.height=5, fig.width=9}
babynames %>% 
  filter(name == "John" & sex == "M" | 
           name == "Elizabeth" & sex == "F") %>% 
  ggplot(aes(year, prop)) +
  geom_line(aes(color = sex)) +
  geom_point(alpha = 0.4, shape = "cross") +
  geom_text(data = . %>% filter(year == 1965), aes(label = name),
            nudge_y = .009) +
  labs(title = "Name Popularity") + 
  theme(legend.position = "none")           
```

Global = `ggplot(aes(_______))`

Override = individual layers: `geom_---(aes(______))`

---
class: middle
```{r echo=TRUE, eval=FALSE}
library(babynames)
babynames %>% 
  ggplot(aes(year, prop)) +  #<<
  geom_line(aes(color = sex)) +  #<<
  geom_point(alpha = 0.4, shape = "cross")  #<< 
```

```{r echo=FALSE, fig.height=5, fig.width=9}
babynames %>% 
  filter(name == "John" & sex == "M" | 
           name == "Elizabeth" & sex == "F") %>% 
  ggplot(aes(year, prop)) +
  geom_line(aes(color = sex)) +
  geom_point(alpha = 0.4, shape = "cross") +
  geom_text(data = . %>% filter(year == 1965), aes(label = name),
            nudge_y = .009) +
  labs(title = "Name Popularity") + 
  theme(legend.position = "none")           
```


.tiny[.footnote[Next slide for actual code]]

---
class: middle

```r
babynames %>% 
  filter(name == "John" & sex == "M" | 
           name == "Elizabeth" & sex == "F") %>% 
  ggplot(aes(year, prop)) +
  geom_line(aes(color = sex)) +
  geom_point(alpha = 0.4, shape = "cross") +
  geom_text(data = . %>% filter(year == 1965), aes(label = name),
            nudge_y = .009) +
  labs(title = "Name Popularity") + 
  theme(legend.position = "none")           
```


---
class: middle
## Facet wrap


```{r echo=TRUE, fig.height=5, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ class)  #<<
```


---
class: middle
## Labels

```{r fig.height=5, fig.width=9}
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  labs(title = "Automobile mileage",               #<<
       x = "Displacement", y = "Highway mpg",      #<<
       caption = "Source:  ggplot2::mpg")          #<<
```

https://ggplot2.tidyverse.org/reference/labs.html

---
## Scales

https://ggplot2.tidyverse.org/reference/index.html#section-scales

---
## Themes

https://ggplot2.tidyverse.org/reference/ggtheme.html


---
## Effective Visualization

[Designing effective visualizations](https://introds.org/slides/w4_d1-effective-dataviz/w4_d1-effective-dataviz.html#26) by Dr. Mine Çetinkaya-Rundel - _Introduction to Data Science_ https://introds.org

[Data Visualization: A Practical Introduction](https://socviz.co/lookatdata.html) Kieran Healy



---
## Reinforce your learning 

On your own...

Interactive Exercises from [RStudio Primers -- Visuzliation](https://rstudio.cloud/learn/primers/3)

[Angela Zoss's code exercises](https://github.com/amzoss/ggplot2-DF19) 

---

# John Little

&nbsp;

.pull-left[.full-width[.content-box-green[
### Data Science Librarian

- https://johnlittle.info/

- https://github.com/libjohn
]]

### Rfun host...
You can make **Rfun** with our resources for R and data science analytics.  See the [R we having fun yet‽](https://rfun.library.duke.edu/) resource pages.
]

.pull-right[.content-box-grey[
### Duke University

**Center for Data & Visualization Sciences**

- https://library.duke.edu/data/
- askData@Duke.edu
- The /Edge, Bostock (1st Floor)
- [Past Workshops](https://library.duke.edu/data/news/past-workshops)
- [Guides & Tutorials](https://library.duke.edu/data/tutorials)

]]




---
class: center, middle
## Shareable 

Data, presentation, and handouts 

<span class="cc">
<!-- Key: a: SA, b: BY, c: CC Circle, d: ND, n: NC, m: Sampling, s: Share, r: Remix, C: CC Full Logo -->
C bn
</span>


[CC BY-NC license](https://creativecommons.org/licenses/by-nc/4.0/)


